 

### 2021-06-13 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

Biophysically, greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016]. Accordingly, in temperate species, heat stress caused due to increase in temperature and VPD even under adequete water caused greater tree hydraulic stress and altered soil-water dynamics than milder drought [@ruehrWaterAvailabilityDominant2016]. However, soil water loss during drought effects larger trees more than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [Norway spruce study, bennett; @mcgregor_tree_nodate]. Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in tropical canopies where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].

@@

**There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].
This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. `r fig_leaf_T` **???**). 
However, evidence in tropical forests also shows that higher nighttime temperatures is associated with decelarating tree growth rates predominantly in smaller trees, but not restricted to, perhaps because increase in nighttime metabolic respiration could supress growth [*Clark 2010; 2013; Feeley et al. 2007*]. It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures. Further, interacting biophysical and biological stressors also show that stress sensitivity and ability to tolerate number of stressors increases with height. Partly because NSC reserves are greater in taller trees due to available light than in buffered trees that are light and NSC limited, therefore making buffered small trees more sensitive to stress [@niinemetsResponsesForestTrees2010a]. In other cases, warming is associated with taller understory trees with lower LMA that complete their lifecycle under closed-canopy temperate forests [@maesPlantFunctionalTrait2020].

However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. `r fig_leaf_T`).

@@

### canopy disturbance and warming

@@

Warming shifts forest composition, from leaf metabolism to whole forest structure. At a leaf metabolism level, fluxnet global analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021]. This can also potentially alter species thermoregulation [@sastry_leaf_2018]. Temperature driven photosynthesis limitation is also reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020]. Feedback loops from loss of big trees through natural disturbance or anthropogenic (e.g.@senfCanopyMortalityHas2018a, 2020;  @mcdowellMechanismsPlantSurvival2008; Brienen et al. 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a] would increase canopy roughness, and reductions in canopy cover is associated with reduced canopy structural complexity [@senfMappingForestDisturbance2020]. Leading to altered biophysical gradient with local heating effects [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].

In the face of global climate change, resiliency among forests are also observed, such as African rainforests [Amy Bennett's new paper] that are more resilient to warming and drought than other forests. Increasing temperature acclimation among tropical and boreal species [martijn's papers; Danielle Way's papers; Atkin's papers] and reversible recovery of canopy heat stress [Smith et al. 2020] are other mechanisms [for eg. *REF*] that will help cope with warming.

>>>>

Biophysical conditions can alter tree-water dynamics and growth patterns. Greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016]. Along with greater VPD, increase in temperature even under adequete water can cause greater tree hydraulic stress due to altered soil-water dynamics than milder drought, in temperate species. Recovery time from heat stress varies between species hydraulic traits [@ruehrWaterAvailabilityDominant2016; @parkwilliamsTemperaturePotentDriver2013]. Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979]. Altered precipitation regimes causing drought is more detrimental to larger trees more than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [Norway spruce study, @bennettLargerTreesSuffer2015; @mcgregor_tree_nodate; @bartholomewSmallTropicalForest2020]. However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. `r fig_leaf_T`).

Tolerance to multiple biophysical and biological stressors can differ among tree size classes, which has shown to increase with tree height and light due to greater non-structural carbon (NSC) reserves, with smaller understory trees being more sensitive to stress [@niinemetsResponsesForestTrees2010a]. This can also alter tree growth patterns. Particularly higher temperatures has shown to primarily reduce tree growth by altering water availability, in mesic temperate and subtropical eucalytus forests [@bowmanWarmerWorldWill2014]. Evidence in tropical forests also shows that higher nighttime temperatures is associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees, perhaps due to increase in nighttime metabolic respiration that could supresses growth [*Clark 2010; 2013; Feeley et al. 2007*]. .

**There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].
This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. `r fig_leaf_T` **???**). 
 It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures. In other cases, warming is associated with taller understory trees with lower LMA that complete their lifecycle under closed-canopy temperate forests [@maesPlantFunctionalTrait2020].

@@

@@

### canopy disturbance

@@

###warming
Warming shifts forest composition, from leaf metabolism to whole forest structure. At a leaf metabolism level, fluxnet global analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021]. However, taken into account light-induced lower leaf respiration in the day, partionationing of C losses differ diurnally and seasonally across ecosytems [Keenan et al. 2019]. However, nitrogen that is particularly essential for plant metabolic processes has also shown to decline in forests with warming [Pamela templer et al.].  Temperature driven photosynthesis limitation is also reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].

Feedback loops from loss of big trees through natural disturbance or anthropogenic (e.g.@senfCanopyMortalityHas2018a, 2020;  @mcdowellMechanismsPlantSurvival2008; Brienen et al. 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a] would increase canopy roughness, and reductions in canopy cover is associated with reduced canopy structural complexity [@senfMappingForestDisturbance2020]. Leading to altered biophysical gradient with local heating effects [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].

In the face of global climate change, resiliency among forests are also observed, such as African rainforests [Amy Bennett's new paper] that are more resilient to warming and drought than other forests. Increasing temperature acclimation among tropical and boreal species [martijn's papers; Danielle Way's papers; Atkin's papers] and reversible recovery of canopy heat stress [Smith et al. 2020] are other mechanisms [for eg. *REF*] that will help forests cope with warming.
